Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-07 conclusions.tex

In this paper, we have presented the algorithms underlying \enzo: an
open-source adaptive mesh refinement code designed for
self-gravitating compressible fluid dynamics, including the effects of
magnetic fields, radiation transport, and a variety of microphysical
and subgrid processes.  In addition, we have described the \enzo\ code
development process, have shown the outputs of a set of representative
test problems, and have provided information about \enzo's performance
and parallel scaling on recent supercomputing platforms.  The \enzo\
code, its test suite, and all of the scripts used to generate plots
and figures for this paper are open source and are available at the
\enzo\ website, \url{http://enzo-project.org}.  Furthermore, the \texttt{yt} toolkit,
which is designed to analyze \enzo\ data (as well as data from a wide
variety of other simulation tools), can be found at its website,
\url{http://yt-project.org}.  Both of these codes have active user and
developer communities, extensive documentation and user support, and
strong mechanisms for users to contribute their changes and fixes to
the codebase.

@@

\begin{itemize}
\item The creation of a hybrid-parallel version of \enzo, combining
MPI for communication between nodes of a supercomputer and OpenMP for
thread-based parallelism within a node.  This will reduce on-node
memory usage and overall scaling behavior.
\item The restructuring of \enzo's treatment of particles to
accommodate a wider range of ``active'' particles that can easily
interact with each other and with multiple grids, and include sink,
source, and particle creation, destruction, splitting, and merging functionality.
\item A new HYPRE-based AMR gravity solver that is faster, more
accurate, and more scalable than the current multigrid solver.
\item New infrastructure for problem initialization, enabling
  users to more quickly and easily create new types of simulations.
\end{itemize}

With the continual rapid development of computer hardware, it makes sense to review
not only at \enzo's current capabilities, but to look towards its future development 
in view of predicted technological trends. These trends in supercomputing hardware suggest that  
substantial modifications to \enzo's core infrastructure, and very possibly some of 
the core algorithms will be required. More
specifically, the progression involves the usage of specialized
large-core-count, vectorized computing units such as graphics
processing units (GPUs) or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to increase, likely much faster than the
bandwidth between nodes, and will require tremendous reduction in
inter-node (and possibly inter-CPU) communication in order to maintain
code scalability.  Also, much of the current code will need to be
rewritten to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures. We therefore anticipated that 
\enzo\ (or a code that has the capabilities of \enzo, from a user's
point of view) will continue to develop to be usable at the largest scales on
such machines.

>>>>

In this paper, we have presented the algorithms underlying \enzo, an
open-source adaptive mesh refinement code designed for
self-gravitating compressible fluid dynamics, including the effects of
magnetic fields, radiation transport, and a variety of microphysical
and subgrid processes.  In addition, we have described the \enzo\ code
development process, have shown the outputs of a set of representative
test problems, and have provided information about \enzo's performance
and parallel scaling on recent supercomputing platforms.  The \enzo\
code, its test suite, and all of the scripts used to generate plots
and figures for this paper are open source and are available at the
\enzo\ website, \url{http://enzo-project.org}.  Furthermore, the
\texttt{yt} toolkit, which is designed to analyze \enzo\ data (as well
as data from a wide variety of other simulation tools), can be found
at its website, \url{http://yt-project.org}.  Both of these codes have
active user and developer communities, extensive documentation and
user support, and strong mechanisms for users to contribute their
changes and fixes to the codebase.

@@

\begin{itemize}
\item The creation of a hybrid-parallel version of \enzo, combining
MPI for communication between nodes of a supercomputer and OpenMP for
thread-based parallelism within a node.  This will reduce on-node
memory usage and improve overall scaling behavior.
\item The restructuring of \enzo's treatment of particles to
accommodate a wider range of ``active'' particles that can easily
interact with each other and with multiple grids, and include sink,
source, and particle creation, destruction, splitting, and merging functionality.
\item A new HYPRE-based AMR gravity solver that is faster, more
accurate, and more scalable than the current multigrid solver.
\item New infrastructure for problem initialization, enabling
  users to more quickly and easily create new types of simulations.
\end{itemize}

With the continual rapid development of computer hardware, it makes
sense to not only review \enzo's current capabilities, but to look
toward its future development in view of predicted technological
trends. These trends in supercomputing hardware suggest that
substantial modifications to \enzo's core infrastructure, and very
possibly some of the core algorithms, will be required. More
specifically, the progression involves the usage of specialized
large-core-count, vectorized computing units such as graphics
processing units or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to increase, likely much faster than the bandwidth
between nodes, and will require tremendous reduction in inter-node
(and possibly inter-CPU) communication in order to maintain code
scalability.  Also, much of the current code will need to be rewritten
to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures. We therefore anticipated
that \enzo\ (or a code that has the capabilities of \enzo, from a
user's point of view) will continue to be usable at the
largest scales on such machines.

### 2013-06-07 development.tex

Over time, \enzo's development has followed a trajectory toward
openness.  Started as the graduate research project of Greg Bryan at
the University of Illinois, it has subsequently been stewarded by
the Laboratory for Computational Astrophysics (LCA) at the University
of California at San Diego, and has transitioned to a distributed,
completely open, and community-driven project.  Initially, \enzo\ was
versioned using a series of ``snapshots'' of the code base, usually
hand-created by the developing individuals.  These were distributed to
collaborators and colleagues, but the central ``trunk'' of development
was updated primarily by a single person: while patches and technology
were accepted from external developers, the relatively small number of
individuals using the code resulted in a strong centralization of
development.

@@

Following the first public, open source release of \enzo, the code was
migrated to using the Subversion version control system.  This is a
centralized version control system, and the ``stable'' \enzo\ source
code was made globally readable following the \enzo\ 1.5 release.
Access to the primary development tree required a password and login
for each user, and providing upstream changes either required this
password and the granting of write access or a sequence of patches and
manually-created diffs (much like the original development system).
The technical friction of manually contributing patches and
modifications, combined with Subversion's difficulty with tracking
merges, resulted in further fragmentation of the code base.

@@

Currently, \enzo\ is developed using the hosted source control
platform Bitbucket (\url{http://bitbucket.org/}) at
\url{http://bitbucket.org/enzo/}.  There are two mailing lists, one
for usage-focused questions and discussion, and another for
development discussion.  Both of these lists are open and publicly
archived.  Bitbucket provides mechanisms for inspecting source,
hosting branches and forks of the primary source, and for code review.
All proposed source code changes for \enzo\ are subjected to a peer
review process, where experienced developers read, inspect, test, and
provide feedback on source code changes.  All developers, including
long-time \enzo\ contributors and developers, are subject to this
process before their code is included in the primary \enzo\
repository.  By using a remote, hosted system, \enzo\ is now
\textit{completely} open to contributions from the community.
Individuals, who may or may not consider themselves \enzo\ developers,
are free to ``fork'' the \enzo\ code base, develop changes (signed
with their own name), and submit them for review and inclusion.  In
contrast to the centralized, gatekeeper-focused technology used
previously, this enables anyone to contribute changes to be evaluated
for inclusion in the \enzo\ codebase.

While peer review is able to catch many bugs and problems with source
code changes, \enzo\ is also subject to ``answer'' testing.  We have
created a set of parameter files and problem types that exercise the
underlying machinery of \enzo.  These ``test problems'' have
affiliated ``tests,'' which consist of scripts that use \texttt{yt}
\citep{2011ApJS..192....9T} to produce results such as mass
distribution, projections, profiles and so on.  The testing
infrastructure then evaluates whether the variation in the new results
compared to a ``gold standard'' of results has exceeded an acceptable
threshold, typically set to roughly single precision.  Optionally, for
those test problems that are deemed unsafe to change to any precision,
the tests also produce hashes of the outputs; these hashes will only
remain unchanged in the event of bitwise identicality between results.
The results of the gold standard are versioned and stored in Amazon
S3, enabling remote testing to proceed.  While the testing process --
building, running, analyzing and comparing -- is not yet automated
against incoming pull requests, we hope to deploy that functionality
in the future.  The primary challenge is that of compute time; the
tests are organized into multiple categories, including by the
expected run time, but the full suite of tests can take several days
to run.

>>>>

Over time, \enzo's development has followed a trajectory toward
openness.  Started as the graduate research project of Greg Bryan at
the University of Illinois, it was subsequently stewarded by the
Laboratory for Computational Astrophysics (LCA) at the University of
California at San Diego, and has transitioned to a distributed,
completely open, and community-driven project.  Initially, \enzo\ was
versioned using a series of ``snapshots'' of the code base, usually
hand-created by the individuals doing development work.  These were
distributed to collaborators and colleagues, but the central ``trunk''
of development was updated primarily by a single person: while patches
and technology were accepted from external developers, the relatively
small number of individuals using the code resulted in a strong
centralization of development.

@@

Following the first public, open source release of \enzo, the code was
migrated to the Subversion version control system.  This is a
centralized version control system, and the ``stable'' \enzo\ source
code was made globally readable following the \enzo\ 1.5 release.
Access to the primary development tree required a password and login
for each user, and providing upstream changes either required this
password and the granting of write access or a sequence of patches and
manually-created diffs (much like the original development system).
The technical friction of manually contributing patches and
modifications, combined with Subversion's difficulty with tracking
merges, resulted in further fragmentation of the code base.

@@

Currently, \enzo\ is developed using the hosted source control
platform Bitbucket (\url{http://bitbucket.org/}) at
\url{http://bitbucket.org/enzo/}.  There are two mailing lists, one
for usage-focused questions and discussion, and another for
development discussion.  Both of these lists are open and publicly
archived.  Bitbucket provides mechanisms for inspecting source code,
hosting branches and forks of the primary source, and for code review.
All proposed source code changes for \enzo\ are subjected to a peer
review process, where experienced developers read, inspect, test, and
provide feedback on source code changes.  All developers, including
long-time \enzo\ contributors and developers, are subject to this
process before their code is included in the primary \enzo\
repository.  By using a remote, hosted system, \enzo\ is now
\textit{completely} open to contributions from the community.
Individuals, who may or may not consider themselves \enzo\ developers,
are free to ``fork'' the \enzo\ code base, develop changes (signed
with their own name), and submit them for review and inclusion.  In
contrast to the centralized, gatekeeper-focused technology used
previously, this enables anyone to contribute changes to be evaluated
for inclusion in the \enzo\ codebase.

While peer review is able to catch many bugs and problems with source
code changes, \enzo\ is also subject to ``answer'' testing (described
in detail in Section~\ref{sec.tests.suite}).  We have created a set of
parameter files and problem types that exercise the underlying
machinery of \enzo.  These ``test problems'' have affiliated
``tests,'' which consist of scripts that use \texttt{yt}
\citep{2011ApJS..192....9T} to produce results such as mass
distribution, projections, profiles and so on.  The testing
infrastructure then evaluates whether the variation in the new results
compared to a ``gold standard'' of results has exceeded an acceptable
threshold, typically set to roughly the level of roundoff error in
single-precision floating point arithmetic.  Optionally, for those
test problems that are deemed unsafe to change to any precision, the
tests also produce hashes of the outputs; these hashes will only
remain unchanged in the event of bitwise identicality between results.
The results of the gold standard are versioned and stored in Amazon
S3, enabling remote testing to proceed.  While the testing process --
building, running, analyzing and comparing -- is not yet automated
against incoming pull requests, we hope to deploy that functionality
in the future.  The primary challenge is that of compute time; the
tests are organized into multiple categories, including by the
expected run time, but the full suite of tests can take several days
to run.

### 2013-06-07 introduction.tex

% EJT: in code comments point to changes

@@

Due to the high spatial and temporal dynamical ranges involved, astrophysics and cosmological phenomena present a taxing challenge for simulators. To tackle such situations, a number of numerical techniques have been developed which can be broadly split into the gridless, fully Lagrangian methods and grid-based, Eulerian schemes. The most commonly used is an example of the first type known as Smoothed Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved much success, particularly in regimes dominated by gravity.
% I felt metioning SPH before anything else was a little confusing for the article's focus so I added an earlier mention of Eulerian. 
%where systems evolve over large spatial and temporal dynamic ranges (cut because we've already said that in the first sentense - EJT)
However, its development to include an increasing number of sort-after physical processes is still at a relatively early stage when compared with the effort put into the latter type of Eulerian grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.
% is that what was meant by `` Its development, however, is still at a relatively early stage''? Since SPH is older than AMR, I found this a little confusing as written.

Despite this invested expertise, the Eulerian solvers in their original form have a serious drawback: they do not provide an easy method of adaptively increasing the spatial and temporal resolution in small volumes of the simulation. Such flexibility is essential for following physical processes such as gravitational instability. 
A solution to this was first proposed by \citet{Berger89} in the Computational Fluid Dynamics (CFD) community, and became known as Structured Adaptive Mesh Refinement (SAMR). The principal is to adaptively add and modify additional, finer meshes (``grids'') over regions that require higher resolution. In addition, it is possible to add other more advanced physics including --for the implementation in the astrophysics code presented in this paper-- comoving coordinates, self-gravity, radiative cooling, chemistry, heat conduction, collisionless fluids, magnetohydrodynamics, radiation transport, star formation and a range of other physical effects.
% I was concerned the original text sounded a little like this was a brand new technique in a brand new code (despite the references below) which would be confusing to a reader familiar with the field and in particular, Enzo.

There have been a number of numerical methods described in the astronomical literature which contain elements of SAMR or have a similar aim. For example, the N-body solver developed by \citet{Villumsen89} used non-adaptive meshing to increase the resolution in pre-selected regions. This static approach was later used extensively when applied to hydrodynamics \citep[e.g.,][]{Ruffert94, Anninos94}. Adding adaptivity is a more recent enhancement, and there are now a number of N-body codes that possess this feature, both with and without  hydrodynamics  \citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97, Truelove98, flash_method, MLAPM01,  Yahagi01, RAMSES, Quilis04, Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}. Of these, perhaps the most comparable and widely used are FLASH \citep{flash_method}, which uses grid blocks of fixed size, and RAMSES \citep{RAMSES} and ART \citep{ART97}, both of which refine individual cells. It is also possible to deform the grid to obtain high resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a few grid codes have adopted an unstructured approach based on a moving Voronoi mesh \citep{Arepo10, Tess11}.

@@

% 'finally' here sounded strange, since this is really the key point
Numerical simulations of astronomical phenomena now play a key role, along with observations and analytic theory, in pushing forward our understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01, DecadalSurvey10}.   But along with this role comes responsibility.  We believe that those developing simulation tools must fulfill two key obligations: the first is to make those tools available to the community as a whole, much in the way that astronomical data is now regularly made publicly available.  The second is to document, test and refine those methods so that they can be critically evaluated and expanded by others.  This paper represents one part of our attempt to meet our obligations with respect to the \enzo\ code.

The structure of the code is as follows.  In Section~\ref{sec.overview}, we first provide a top-level overview of the code method and structure.  This is designed to give a first broad-brush picture of the equations we aim to solve and the methods used to solve them.  Next, in Sections~\ref{sec.amr} through~\ref{sec.num.analysis}, we describe the methods we use in detail, reserving some of the longer descriptions of particular components for the appendix in order not to interrupt the flow of the paper.  The \enzo\ testing framework and code tests are described in Section~\ref{sec.tests}.  The parallelism strategy and scaling results are described in Section~\ref{sec.parallel}.  Finally, we discuss the code's development methodology (which is as far as we know unique in the astrophysics community) in Section~\ref{sec.development}.

>>>>

@@

Due to the high spatial and temporal dynamical ranges involved,
astrophysics and cosmological phenomena present a taxing challenge for
simulators. To tackle such situations, a number of numerical
techniques have been developed that can be broadly split into the
gridless, Lagrangian methods and grid-based, Eulerian schemes. The
most commonly used is an example of the first type known as Smoothed
Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved
much success, particularly in regimes dominated by gravity. However,
its development to include an increasing number of sought-after
physical processes is still at a relatively early stage when compared
with the effort put into the latter type of Eulerian grid-based
hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997,
Woodward84}.

Despite this invested expertise, the Eulerian solvers in their
original form have a serious drawback: they do not provide an easy
method of adaptively increasing the spatial and temporal resolution in
small volumes of the simulation. Such flexibility is essential for
following physical processes such as gravitational instability. A
solution to this was first proposed by \citet{Berger89} in the
Computational Fluid Dynamics (CFD) community, and became known as
Structured Adaptive Mesh Refinement (SAMR). The principle is to
adaptively add and modify additional, finer meshes (``grids'') over
regions that require higher resolution. In addition, it is possible to
add other, more advanced physics including -- for the AMR
implementation in the astrophysics code presented in this paper --
comoving coordinates, self-gravity, radiative cooling, chemistry, heat
conduction, collisionless fluids, magnetohydrodynamics, radiation
transport, star formation and a range of other physical effects.

There have been a number of numerical methods described in the
astronomical literature that contain elements of SAMR or have a
similar aim. For example, the N-body solver developed by
\citet{Villumsen89} used non-adaptive meshing to increase the
resolution in pre-selected regions. This static approach was later
used extensively when applied to hydrodynamics
\citep[e.g.,][]{Ruffert94, Anninos94}. Adding adaptivity is a more
recent enhancement, and there are now a number of N-body codes that
possess this feature, both with and without hydrodynamics
\citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97,
Truelove98, flash_method, MLAPM01, Yahagi01, RAMSES, Quilis04,
Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}. Of these,
perhaps the most comparable and widely used are FLASH
\citep{flash_method}, which uses grid blocks of fixed size, and RAMSES
\citep{RAMSES} and ART \citep{ART97}, both of which refine individual
cells. It is also possible to deform the grid to obtain high
resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a
few grid codes have adopted an unstructured approach based on a moving
Voronoi mesh \citep{Arepo10, Tess11}.

@@

Numerical simulations of astronomical phenomena now play a key role,
along with observations and analytic theory, in pushing forward our
understanding of the cosmos \citep[e.g.,][]{DecadalSurvey01,
DecadalSurvey10}.  But along with this role comes responsibility.  We
believe that those developing simulation tools must fulfill two key
obligations: the first is to make those tools available to the
community as a whole, much in the way that astronomical data is now
regularly made publicly available.  The second is to document, test
and refine those methods so that they can be critically evaluated and
expanded upon by others.  This paper represents one component of our
attempt to meet these obligations with respect to the \enzo\ code.

The structure of this paper is as follows.  In
Section~\ref{sec.overview}, we first provide a top-level overview of
the code method and structure.  This is designed to give a broad-brush
picture of the equations solved by \enzo\ and the methods used to
solve them.  Next, in Sections~\ref{sec.amr}
through~\ref{sec.num.analysis}, we describe the methods we use in
detail, reserving some of the longer descriptions of particular
components for the appendix in order to not interrupt the flow of the
paper.  The \enzo\ testing framework and code tests are described in
Section~\ref{sec.tests}.  The parallelism strategy and scaling results
are described in Section~\ref{sec.parallel}.  Finally, we discuss the
code's development methodology (which is, as far as we know, unique in
the astrophysics community) in Section~\ref{sec.development}.

### 2013-06-07 numerical-chemistry.tex

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by Abel et al. and Anninos et al. \citep{abel97,anninos97}, but have
been extended since then with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  Enzo
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that faster than those of the gas dynamics.

@@

\end{tabular}
\caption[]{Collisional processes solved in the Enzo nonequilibrium
primordial chemistry routines.}
\label{table.collisional}
\end{center}
\end{table}

\begin{table}
\begin{center}
{\bfseries Radiative Processes}\\[1ex]
\begin{tabular}{llllllll}
(32) & H & + & $\gamma$ & $\rightarrow$ & H$^+$ & + & e$^-$ \\
(33) & He & + & $\gamma$ & $\rightarrow$ & He$^+$ & + & e$^-$ \\
(34) & He$^+$ & + & $\gamma$ & $\rightarrow$ & He$^{++}$ & + & e$^-$ \\
(35) & H$^-$ & + & $\gamma$ & $\rightarrow$ & H & + & e$^-$ \\
(36) & H$_2$ & + & $\gamma$ & $\rightarrow$ & H$_2^+$ & + & e$^-$ \\
(37) & H$_2^+$ & + & $\gamma$ & $\rightarrow$ & H & + & H$^+$ \\
(38) & H$_2^+$ & + & $\gamma$ & $\rightarrow$ & 2H$^+$ & + & e$^-$ \\
(39) & H$_2$ & + & $\gamma$ & $\rightarrow$ & H$_2^*$ & $\rightarrow$ & 2H \\
(40) & H$_2$ & + & $\gamma$ & $\rightarrow$ & 2H &  & 
\end{tabular}
\caption[]{Radiative processes solved in the Enzo nonequilibrium
primordial chemistry routines. \red{Missing rates for HD destruction/ionization?}}
\label{table.radiative}
\end{center}
\end{table}

>>>>

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by Abel et al. and Anninos et al. \citep{abel97,anninos97}, but have
been extended since then with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation; see Section~\ref{sec.num.cooling} for
more details.  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., de-coupled and operator split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that faster than those of the gas dynamics.

@@

\end{tabular}
\caption[]{Collisional processes solved in the \enzo\ nonequilibrium
primordial chemistry routines.}
\label{table.collisional}
\end{center}
\end{table}

\begin{table}
\begin{center}
{\bfseries Radiative Processes}\\[1ex]
\begin{tabular}{llllllll}
(32) & H & + & $\gamma$ & $\rightarrow$ & H$^+$ & + & e$^-$ \\
(33) & He & + & $\gamma$ & $\rightarrow$ & He$^+$ & + & e$^-$ \\
(34) & He$^+$ & + & $\gamma$ & $\rightarrow$ & He$^{++}$ & + & e$^-$ \\
(35) & H$^-$ & + & $\gamma$ & $\rightarrow$ & H & + & e$^-$ \\
(36) & H$_2$ & + & $\gamma$ & $\rightarrow$ & H$_2^+$ & + & e$^-$ \\
(37) & H$_2^+$ & + & $\gamma$ & $\rightarrow$ & H & + & H$^+$ \\
(38) & H$_2^+$ & + & $\gamma$ & $\rightarrow$ & 2H$^+$ & + & e$^-$ \\
(39) & H$_2$ & + & $\gamma$ & $\rightarrow$ & H$_2^*$ & $\rightarrow$ & 2H \\
(40) & H$_2$ & + & $\gamma$ & $\rightarrow$ & 2H &  & 
\end{tabular}
\caption[]{Radiative processes solved in the \enzo\ nonequilibrium
primordial chemistry routines. \red{Missing rates for HD destruction/ionization?}}
\label{table.radiative}
\end{center}
\end{table}

### 2013-06-07 numerical-mhd-ct.tex

Divergence-free AMR is somewhat more complex than the AMR employed
elsewhere in \enzo.  First, the interpolation must be constrained to be
divergence free.  Thus, all three face-centered field components are
interpolated in concert.  Second, any magnetic field information from
the previous time step must be included in the interpolation, making the
interpolation more complex than the simple parent-child relation used
for other fields.  Third, the flux correction involves more possible
grid relations than traditional AMR.  In order to circumvent this last
complexity, the electric field is projected from fine grids to parent
grids (rather than the magnetic field), and is then used to re-update
the parent magnetic field.  This is described in detail in
\citet{Balsara99} and \citet{Collins10}.

>>>>

Divergence-free AMR is somewhat more complex than the AMR employed
elsewhere in \enzo.  First, the interpolation must be constrained to be
divergence free.  Thus, all three face-centered field components are
interpolated in concert.  Second, any magnetic field information from
the previous timestep must be included in the interpolation, making the
interpolation more complex than the simple parent-child relation used
for other fields.  Third, the flux correction involves more possible
grid relations than traditional AMR.  In order to circumvent this last
complexity, the electric field is projected from fine grids to parent
grids (rather than the magnetic field), and is then used to re-update
the parent magnetic field.  This is described in detail in
\citet{Balsara99} and \citet{Collins10}.

### 2013-06-07 numerical-starformation.tex
Edits to the overview section. Trying to unify notations, so several other sections are also affected.

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, every cell at the
locally highest level of refinement are examined at every timestep to
see if they meet the critiera for star formation, and star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
%\end{equation} 
%\begin{equation}
\nabla \cdot \vec{v}_b & < & 0, \\
\label{cencont}
%\end{equation}
%\begin{equation}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
%\end{equation}
%\begin{equation}
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
%$\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
%the mean baryon density in the simulation, 
$\eta$ is the user-defined
overdensity threshold, 
%$\vec{v}$ is the velocity of the gas in the
%cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
%\rho_{b} + \rho_{d}$, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

Algorithmically, the star formation events in the Schmidt-law algorithm are
assumed to occur once every global time step (with the constraint $\Delta t_0 \leq
10^7$ years).
In cells where star formation is determined to occur (i.e. $\rho_{\rm
  gas} \geq \rho_{\rm SF}$), star particles with a mass of $m_* =
\dot{\rho}_* V_{\rm cell} \Delta t_0$ (where $V_{\rm cell}$ is the
volume of the mesh cell) are assumed to form instantaneously in a
manner similar to that described in Section~\ref{sec:starform_cen}.
The \enzo\ implementation of this algorithm is similar, except that
instead of forming stars only at the root grid time step, we allow
stars to form at the time step of the highest level of resolution at
any particular point in space.  As can be seen from the equation for
$m_*$ above, this can result in very small stellar masses.  To avoid
memory and processor time issues related to having very large numbers
of star particles we impose a threshold mass $M_{\rm *,min}$ such that
a star particle only forms if $m_* \geq M_{\rm *,min}$.  An
appropriate choice of $M_{\rm *,min}$ does not significantly change
the star overall star formation history of a simulation, though it may
delay the onset of star formation in a given cell relative to a
simulation without a particle mass threshold.

@@

\item A converging gas flow ($\nabla \cdot \vec{v} < 0$), and

>>>>

The Cen \& Ostriker method is a heuristic model of star formation on
galactic scales.  This method, first described by \citet{CO1992},
assumes that stars form in substantially overdense, converging, and
gravitationally unstable gas.  Algorithmically, every cell at the
locally highest level of refinement are examined at every timestep to
see if they meet the critiera for star formation, and star particles
are allowed to form in a cell if the following criteria are met:
\begin{eqnarray}
\rho_b/\bar{\rho}_b & \geq & \eta,  \\[2mm]
\label{cendens}
%\end{equation} 
%\begin{equation}
\div \vec{v}_b & < & 0, \\
\label{cencont}
%\end{equation}
%\begin{equation}
t_{\rm cool} & < & t_{\rm dyn} \equiv \sqrt{3 \pi / 32G \rho_{\rm tot}}, \\
%\end{equation}
%\begin{equation}
m_{b} & > & m_{J} \equiv G^{-3/2} \rho_{b}^{-1/2}c^{3} 
\left[ 1 + \frac{\delta\rho_{d}}{\delta\rho_{b}} \right]^{-3/2}
\end{eqnarray}
where 
%$\rho_b$ is the baryon density in that cell, $\bar{\rho_b}$ is
%the mean baryon density in the simulation, 
$\eta$ is the user-defined
overdensity threshold, 
%$\vec{v}$ is the velocity of the gas in the
%cell, $\rho_{d}$ is the dark matter density, $\rho_{\rm tot} =
%\rho_{b} + \rho_{d}$, 
and $m_{b}$ and $m_{j}$ are the baryonic mass in the
cell and the Jeans mass of the cell, and c is the isothermal sound speed
in the cell.  If all of these criteria are met, the mass of a star
particle is calculated as \(m_{*} = m_{b} \frac{ \Delta t}{ t_{\rm
    dyn} } f_{\rm *eff} \), where $f_{\rm *eff}$ is the star formation
efficiency parameter.

@@

Algorithmically, the star formation events in the Schmidt-law algorithm are
assumed to occur once every global timestep (with the constraint $\Delta t_0 \leq
10^7$ years).
In cells where star formation is determined to occur (i.e. $\rho_{\rm
  gas} \geq \rho_{\rm SF}$), star particles with a mass of $m_* =
\dot{\rho}_* V_{\rm cell} \Delta t_0$ (where $V_{\rm cell}$ is the
volume of the mesh cell) are assumed to form instantaneously in a
manner similar to that described in Section~\ref{sec:starform_cen}.
The \enzo\ implementation of this algorithm is similar, except that
instead of forming stars only at the root grid timestep, we allow
stars to form at the timestep of the highest level of resolution at
any particular point in space.  As can be seen from the equation for
$m_*$ above, this can result in very small stellar masses.  To avoid
memory and processor time issues related to having very large numbers
of star particles we impose a threshold mass $M_{\rm *,min}$ such that
a star particle only forms if $m_* \geq M_{\rm *,min}$.  An
appropriate choice of $M_{\rm *,min}$ does not significantly change
the star overall star formation history of a simulation, though it may
delay the onset of star formation in a given cell relative to a
simulation without a particle mass threshold.

@@

\item A converging gas flow ($\div \vec{v} < 0$), and
